---
title: "Data & Cleaning"
subtitle: "Spring 2026"
format: 
  html:
    page-layout: full
    theme: cosmo
    toc: true
    toc-title: "Contents"
    embed-resources: true
    standalone: true
editor: 
  markdown: 
    wrap: 72
---

::::: columns
::: {.column width="40%"}
Dataset: IPUMS Time Use

  * Collection of time diary data
  * American Heritage Time Use Study Extract Builder
  * From 1930 to 2018, harmonized for comparison over time
  
Filters:

  * Day: M-F
  * Household type: married with kids
  * Kids under 18 in household
  * Working-age individuals (25 through 54)
  * College educated
  * Filter out missing/dirty values (wage, sex, time use)
  * Year: 2010s
  
Transformations:

  * Convert categorical variables to factors
  * Convert time use variables to hours per day
  * Combine time spent on child care and domestic work into one "unpaid domestic work" variable
:::
::: {.column width="60%"}
```{r codebook, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

# define codebook
codebook <- tibble::tribble(
  ~Variable, ~Description, ~Type,
  "WAGELM", "Monthly employment income", "Numerical",
  "SEX", "Individual's sex (female=1, male=0)", "Binary",
  "YEAR", "Year of the diary entry", "Numerical",
  "HHID", "Household type (1=married with child)", "Categorical",
  "UNDER18", "Number of children under 18 in the household", "Numerical",
  "AGEYNGST", "Age of youngest child", "Numerical",
  "EDUC", "Education (5=college graduate, 6=post college)", "Categorical",
  "AGE", "Age of participant", "Numerical",
  "ACT_WORK", "Daily time spent on paid work", "Numerical",
  "ACT_UNPAID", "Daily time spent on unpaid domestic work", "Numerical",
  "ACT_CHCARE", "Daily time spent on child care", "Numerical",
  "ACT_CIVIC", "Daily time spent on civic, voluntary, and religious activities", "Numerical",
  "ACT_EDUCA", "Daily time spent on education", "Numerical",
  "ACT_LEISURE", "Daily time spent on free time (leisure)", "Numerical",
  "ACT_PCARE", "Daily time spent on personal care", "Numerical",
  "ACT_TRAVEL", "Daily time spent on travel", "Numerical"
)

# print codebook aesthetically
codebook %>%
  kable(format = "html") %>%
  kable_styling(font_size = 18, full_width = FALSE, position = "center") %>%
  column_spec(1, width = "8em") %>%
  column_spec(2, width = "25em")
```
:::
:::::

```{r load_packages, include=FALSE, warning=FALSE, message=FALSE}
# R packages
library(ipumsr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(reticulate)
library(gt)
```

```{r load_data, include=FALSE, warning=FALSE, message=FALSE}
# load in data
ddi <- read_ipums_ddi("data/ahtus_00002.xml")
data <- read_ipums_micro(ddi)
```

```{r filtering_vis1-4, include=FALSE}
# initial filtering for EDA visuals 1-4
filtered_data <- data %>%
  filter(
    HHTYPE == 1, # filter to married with kids
                 # isolating a group where childcare is relevant
                 # avoid comparing people who don’t face childcare constraints
    UNDER18 > 0, # filter to individuals with kids under age of 18
    AGE >= 22 & AGE <= 40, # between the ages of 22 and 40
    EDUC %in% c(5, 6), # college educated
    AGEYNGST >= 0 & AGEYNGST <= 18, # youngest child is between age 0 and 18
    SEX %in% c(1, 2), # male or female
    !is.na(AGE), 
    !is.na(EDUC)
  ) %>%
  mutate(
    sex_label = ifelse(SEX == 1, "Male", "Female"),
    ACT_UNPAID = ACT_CHCARE + ACT_UNDOM
  )
```


```{r filtering_vis5_reg, include=FALSE}

# filtering and turning sex and full-time into binary variables
datareg <- filtered_data %>%
  
  filter(
    (EDUC == 5 | EDUC == 6), # filter to just college educated individuals
    # FULLTIME == 1, # filter to just full-time workers
    WAGELM > 0, # remove non-earners
    ACT_CHCARE >= 0, # filter out missing or unclean data
    ACT_PCARE >= 0 # filter out missing or unclean data
  ) %>% 
  
  mutate(
    across(starts_with("ACT_"), ~ .x / 60), # convert all time use vars to hours
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME), 
    log_wage = log(WAGELM + 1) # add column for log wage in case it becomes interesting
  )
  
  
# filter to just 2010s data
data2010s <- datareg %>% 
  filter(SAMPLE > 2009)

```

```{r new_filtering, include=FALSE}

# Trying to omit some selection bias with our original filtering because the penalty to motherhood and marriage premium seems to disappear: keeping the age constraint, 
# filtering and turning sex and full-time into binary variables

filtered2 <- data %>%
  filter(
    DIARYDAY %in% 2:6, # filter to M-F responses
    HHTYPE == 1, # filter to married with kids
                 # isolating a group where childcare is relevant
                 # avoid comparing people who don’t face childcare constraints
    UNDER18 > 0, # filter to individuals with kids under age of 18
    AGE >= 25 & AGE <= 54, # between the ages of 25 and 54
    EDUC %in% c(5, 6), # college educated
    AGEYNGST >= 0 & AGEYNGST <= 18, # youngest child is between age 0 and 18
    SEX %in% c(1, 2), # male or female
    WAGELM > 0 | WAGELM == 0,
    ACT_CHCARE >= 0, # filter out missing or unclean data
    ACT_PCARE >= 0, # filter out missing or unclean data
    !is.na(AGE), 
    !is.na(EDUC)
  ) %>%
  mutate(
    sex_label = ifelse(SEX == 1, "Male", "Female"),
    across(starts_with("ACT_"), ~ .x / 60), # convert all time use vars to hours
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME), 
    log_wage = log(WAGELM + 1), # add column for log wage in case it becomes interesting
    ACT_UNPAID = ACT_CHCARE + ACT_UNDOM
  ) 

  
# filter to just 2010s data
data2010s2 <- filtered2 %>% 
  filter(SAMPLE > 2009)

```