---
title: "Results & Discussion"
subtitle: "Spring 2026"
format: 
  html:
    page-layout: full
    theme: cosmo
    toc: true
    toc-title: "Contents"
    embed-resources: true
    standalone: true
editor: 
  markdown: 
    wrap: 72
---

## Results

```{r load_packages, include=FALSE, warning=FALSE, message=FALSE}
# R packages
library(ipumsr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(reticulate)
library(gt)
```

```{r load_data, include=FALSE, warning=FALSE, message=FALSE}
# load in data
ddi <- read_ipums_ddi("data/ahtus_00002.xml")
data <- read_ipums_micro(ddi)
```

```{r filtering_vis1-4, include=FALSE}
# initial filtering for EDA visuals 1-4
filtered_data <- data %>%
  filter(
    HHTYPE == 1, # filter to married with kids
                 # isolating a group where childcare is relevant
                 # avoid comparing people who don’t face childcare constraints
    UNDER18 > 0, # filter to individuals with kids under age of 18
    AGE >= 22 & AGE <= 40, # between the ages of 22 and 40
    EDUC %in% c(5, 6), # college educated
    AGEYNGST >= 0 & AGEYNGST <= 18, # youngest child is between age 0 and 18
    SEX %in% c(1, 2), # male or female
    !is.na(AGE), 
    !is.na(EDUC)
  ) %>%
  mutate(
    sex_label = ifelse(SEX == 1, "Male", "Female"),
    ACT_UNPAID = ACT_CHCARE + ACT_UNDOM
  )
```

```{r filtering_vis5_reg, include=FALSE}

# filtering and turning sex and full-time into binary variables
datareg <- filtered_data %>%
  
  filter(
    (EDUC == 5 | EDUC == 6), # filter to just college educated individuals
    # FULLTIME == 1, # filter to just full-time workers
    WAGELM > 0, # remove non-earners
    ACT_CHCARE >= 0, # filter out missing or unclean data
    ACT_PCARE >= 0 # filter out missing or unclean data
  ) %>% 
  
  mutate(
    across(starts_with("ACT_"), ~ .x / 60), # convert all time use vars to hours
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME), 
    log_wage = log(WAGELM + 1) # add column for log wage in case it becomes interesting
  )
  
  
# filter to just 2010s data
data2010s <- datareg %>% 
  filter(SAMPLE > 2009)

```

```{r new_filtering, include=FALSE}

# Trying to omit some selection bias with our original filtering because the penalty to motherhood and marriage premium seems to disappear: keeping the age constraint, 
# filtering and turning sex and full-time into binary variables

filtered2 <- data %>%
  filter(
    DIARYDAY %in% 2:6, # filter to M-F responses
    HHTYPE == 1, # filter to married with kids
                 # isolating a group where childcare is relevant
                 # avoid comparing people who don’t face childcare constraints
    UNDER18 > 0, # filter to individuals with kids under age of 18
    AGE >= 25 & AGE <= 54, # between the ages of 25 and 54
    EDUC %in% c(5, 6), # college educated
    AGEYNGST >= 0 & AGEYNGST <= 18, # youngest child is between age 0 and 18
    SEX %in% c(1, 2), # male or female
    WAGELM > 0 | WAGELM == 0,
    ACT_CHCARE >= 0, # filter out missing or unclean data
    ACT_PCARE >= 0, # filter out missing or unclean data
    !is.na(AGE), 
    !is.na(EDUC)
  ) %>%
  mutate(
    sex_label = ifelse(SEX == 1, "Male", "Female"),
    across(starts_with("ACT_"), ~ .x / 60), # convert all time use vars to hours
    female = ifelse(SEX == 2, 1, 0),
    female = factor(female),
    FULLTIME = factor(FULLTIME), 
    log_wage = log(WAGELM + 1), # add column for log wage in case it becomes interesting
    ACT_UNPAID = ACT_CHCARE + ACT_UNDOM
  ) 

  
# filter to just 2010s data
data2010s2 <- filtered2 %>% 
  filter(SAMPLE > 2009)

```

### Regression 1: Paid vs. Unpaid Work

*Motivation*

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# time spent on paid work by gender



# filter to full time workers

data2010s2FT <- data2010s2 %>%
  filter(FULLTIME == 1)



data_long <- data2010s2FT %>%

  pivot_longer(

    cols = c(ACT_WORK, ACT_UNPAID),

    names_to = "work_type",

    values_to = "hours"

  ) %>%

  mutate(

    work_type = recode(work_type,

                       "ACT_WORK" = "Paid Work",

                       "ACT_UNPAID" = "Unpaid Domestic Work"),

    female = factor(female, labels = c("Men", "Women"))

  )



ggplot(data_long, aes(x = work_type, y = hours, fill = female)) +

  geom_boxplot(position = position_dodge(width = 0.8)) +

  labs(

    title = "Time Spent on Paid vs. Unpaid Work by Gender",

    subtitle = "Initial look at gender differences in working time for full-time workers",

    x = "Time Spent on Work",

    y = "Hours",

    fill = "Gender"

  ) +

  theme_minimal()



# INTERPRETATION

# Median paid work quite similar

# Median unpaid work for women higher than 3rd quartile for men

```

Women who work full-time spend a similar amount of time on paid work but
more time on unpaid domestic work.

*Results*

::::: columns
::: {.column width="50%"}
```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(stargazer)

# the time tradeoff between increased unpaid work to paid work, by gender

linreg2 <- lm(ACT_WORK ~ ACT_UNPAID*female + EMPSP + FULLTIME, data2010s2)

stargazer(linreg2, type="text", omit.stat = c("rsq", "ser", "f", "adj.rsq", "n"), omit = c("EMPSP", "FULLTIME"))

```
:::

::: {.column width="50%"}
-   When controlling for $\,C_i\,$ and time spent on unpaid domestic
    work, there is no statistically significant difference between males
    and females' time spent at paid work.

-   When controlling for $\,C_i\,$, each additional hour spent on unpaid
    domestic work is associated with spending

    -   0.92 fewer hours on paid work for males, and

    -   0.79 fewer hours on paid work for females.
:::
:::::

**Takeaway**: There is evidence of societal expectations that women
continue handling domestic work despite similar paid work obligations.

### Regression 2: Wage vs. Unpaid Work

*Motivation*

```{r, echo=FALSE}

ggplot(data2010s2FT, aes(x = factor(female, labels = c("Men", "Women")),

                        y = WAGELM, fill = factor(female))) +

  geom_boxplot() +

  labs(

    title = "Wage Distribution by Gender",

    subtitle = "Initial look at gender differences in wages for full-time workers",

    x = "",

    y = "Monthly Wage",

    fill = "Gender"

  ) +

  theme_minimal() +

  theme(legend.position = "none")

```

Women on average earn less per month than men, but still need to
consider controlling for other variables.

*Results*

::::: columns
::: {.column width="50%"}
```{r, echo=FALSE, warning=FALSE, message=FALSE}

linreg5 <- lm(WAGELM ~ ACT_UNPAID * female + EMPSP + FULLTIME, data2010s2)

stargazer(linreg5, type="text", omit.stat = c("rsq", "ser", "f", "adj.rsq", "n"), omit = c("EMPSP", "FULLTIME"))

```
:::

::: {.column width="50%"}
-   When controlling for $\,C_i\,$ and time spent on unpaid domestic
    work, females on average earn \$1,517.20 less per month than males.

-   When controlling for $\,C_i\,$, each additional hour spent on unpaid
    domestic work is associated with earning

    -   \$149 less per month for males, and

    -   \$20 less per month for females.
:::
:::::

<br>

**Takeaway**: Same societal expectations are showing, where time spent
on unpaid domestic work has a larger negative effect on male earnings
than female earnings.

## Discussion

The empirical results confirm our hypothesis that the time trade-off
differs by gender. For women, the results indicate a double-burden.
Women must balance both spheres of paid and domestic work.

For men, each additional hour of unpaid work results in near
displacement of paid work ($0.92$ hours). In contrast, women only lose
($0.79$ hours) of paid work for every hour of domestic labor. Women are
more likely to squeeze domestic tasks into their existing schedules,
without retreating from the labor market. Thus, the results corroborate
that women face a higher total labor burden.

For men, domestic work significantly disrupts earnings ($-\$149$ per
hour). For women, the wage impact is less severe ($-\$20$ per hour).
This may suggest that men's wages are tied to roles demanding overtime
or undivided focus, so domestic work penalizes earning potential. Women
may self-select into roles that better align with their domestic
responsibilities. Their wages are lower to begin with, reflecting
domestic trade-offs. So, additional domestic work has a negligible
impact on their earnings.

In both models, gender significantly modifies the relationship between
domestic work and economic outcomes. The steeper slopes for men indicate
their labor force participation is more sensitive to domestic work.
Conversely, women's labor force participation is far less reactive.
Women are expected to work the same amount or more, while earning less
on average.

## Limitations

-   Diary data is messy and inconsistent for early time periods
    (pre-2010s).\
-   We recognize the existence of potential selection bias within our
    sample. Are our filters too strict or not strict enough?
    -   Household type is constrained to married couples with kids.
    -   Both spouses are college educated.
-   Limited to male/female binary and heterosexual relationships

## Extensions for Future Work

-   Include different demographic variables in regressions as
    explanatory variables:
    -   Race, Education-Level, Religion, etc.
-   Include alternative time use variables as outcomes:
    -   Exercise, Leisure, Personal Care, etc.
-   Time series exploration to examine how these relationships have
    shifted over time
